---
title: "Miriam_Mouse_dataset_preprocessing"
author: "Dillon Corvino"
date: "28/11/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Mouse 


```

### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)

# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- TRUE
do.plot <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# load packages
library("Seurat")
library("SeuratExtend")

# Load the future package
library(future)

# Set the maximum size of globals that can be exported to 50 GB
options(future.globals.maxSize = 50 * 1024^3)  # 50 GB in bytes
getOption("future.globals.maxSize")


# Ensure ggplot2 is the correct version - requires restarting Rsession and loading packages again
ggplot2_switch_version(desired_version = "3.5.0")

```


## Pre-processing {.tabset}

### Ambient RNA correction
```{r ambient_RNA_correction}

# decided to skip ambient RNA correction for first pass quick look

remove.ambient.rna <- FALSE

# note, if ambient rna removal is done, downstream read in of data needs to be altered to read in the strained data as input

if(remove.ambient.rna){
  
  # Create output dirs
  
  output.dir <- paste0("results/QC/Ambien_RNA_correction/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  save.dir <- paste0("saves/strained_outs/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }  
  
  
  #################################
  # Install and load packages
  #################################
  
  #install.packages("SoupX")
  #BiocManager::install("DropletUtils")
  
  # load packages
  library("SoupX")
  library("DropletUtils")
  
  #################################
  # list all directories in data folder
  #################################
  sequence.runs <- list.dirs(path = "data", recursive = FALSE)
  # capture just the folders with the outs/ from run alignment
  pattern.vec <- c("^data/Doce$", "^data/Vehicle$")
  sequence.runs <- grep(paste0(pattern.vec, collapse = "|"), sequence.runs, value = T)
  sequence.runs
  
  
  # iterate over sequence runs, estimate and correct for ambient RNA and save output
  
  for(i in seq_along(sequence.runs)){
    
    # input dir
    
    tmpDir <- sequence.runs[i]
    
    # extract sample info
    run.id <- gsub("data/", "", sequence.runs[i])

    # name the output dir
    save.dir.name <- paste0(save.dir, run.id, "_strainedCounts/")
    
    # If dir exists and therefore output do not recalculate this output / skip to next sample
    if(dir.exists(save.dir.name)){
      print(paste0("Directory ", save.dir.name, " already exists, not recomputing this sample"))
      next()}else{
        print(paste0("Computing for sample ", run.id))
        
        # create output dir
        dir.create(save.dir.name, 
               recursive = T)
      }
    
    
    # load table of cells (toc) and table of droplets (tod)
    toc = Seurat::Read10X(file.path(tmpDir, "filtered_feature_bc_matrix"))
    tod = Seurat::Read10X(file.path(tmpDir, "raw_feature_bc_matrix"))
    
    
    # create a soup channel with gene expression tables
    #sc = SoupChannel(tod$`Gene Expression`, toc$`Gene Expression`) # if multiple assay types
    sc = SoupChannel(tod, toc)
    
    # load cellranger calculated UMAP projections
    meta.dr <- read.csv(file.path(tmpDir, "analysis/umap/gene_expression_2_components/projection.csv"))
    
    # format meta.dr
    rownames(meta.dr) <- meta.dr$Barcode
    meta.dr$Barcode <- NULL
    
    # sanity check that barcodes are in correct order
    if(sum(rownames(sc$metaData) != rownames(meta.dr)) > 0){
      stop("Error: Row names of sc$metaData do not match row names of meta.dr")
    } 
    
    # add dim reduction to soup channel object
    sc = setDR(sc, DR = meta.dr, reductName = c("UMAP_1", "UMAP_2"))
    
    # load cellranger calculated cluster identities 
    meta.clust <- read.csv(file.path(tmpDir, "analysis/clustering/gene_expression_kmeans_10_clusters/clusters.csv"))
    
    # add cluster info to soup channel object
    sc = setClusters(sc, setNames(meta.clust$Cluster, meta.clust$Barcode))
    
    ##################
    # Estimate rho
    ##################
    # note default params are optimised for fresh (not nuclear) single cell experiments 
    # if processing nuclear scRNAseq, consider adjusting prior Rho default estimates
    sc = autoEstCont(sc, forceAccept = TRUE)
    
    dev.copy(pdf, paste0(output.dir, run.id, "_Rho_estimation.pdf"))
    dev.off()
    
    ##################
    # adjust the data
    ##################
    sc.out <- adjustCounts(sc)
    
    ##################
    # save data to file
    ##################
    
    DropletUtils:::write10xCounts(paste0(save.dir.name), sc.out)
    
    ####################################
    # quick look at correction performed 
    ####################################
    
    # what is the most corrected 
    cntSoggy = rowSums(sc$toc > 0)
    cntStrained = rowSums(sc.out > 0)
    mostZeroed = tail(sort((cntSoggy - cntStrained)/cntSoggy), n = 10)
    print("Most zeroed genes")
    print(mostZeroed)
    
    # genes with quantitative difference
    print("Genes with quantitative differences")
    print(
      tail(sort(rowSums(sc$toc > sc.out)/rowSums(sc$toc > 0)), n = 20)
    )
    
    # clean environment
    rm(list = c("sc", "tod", "toc", "meta.dr", "meta.clust", "mostZeroed", "sc.out"))
    gc()
  }
  
}
# Misc info 
# various tools to visualise and inspect ambient RNA removal and detection
# plotMarkerDistribution(sc.temp)
# 
# # identify most zeroed genes
# cntSoggy = rowSums(sc.temp$toc > 0)
# cntStrained = rowSums(sc.out > 0)
# mostZeroed = tail(sort((cntSoggy - cntStrained)/cntSoggy), n = 20)
# mostZeroed
# 
# tail(sort(rowSums(sc.temp$toc > sc.out)/rowSums(sc.temp$toc > 0)), n = 20)
# 
# 
# plotChangeMap(sc.temp, sc.out, "Il1b")
# 
```


### Load dataset
```{r load_dataset}

if(long.compute){
  
# read in filtered matrix data or strained matrix if ambient RNA is removed
      PTEN <- Seurat::Read10X("data/PTEN/filtered_feature_bc_matrix")
      PTEN_CDCP1 <- Seurat::Read10X("data/PTEN_CDCP1/filtered_feature_bc_matrix")

  # create seurat objects
    seurat.pten <- CreateSeuratObject(counts = PTEN,
                                      project = "PTEN")
  
    seurat.pten.cdcp1 <- CreateSeuratObject(counts = PTEN_CDCP1,
                                      project = "PTEN_CDCP1")
    
    # merge seurat objects
    seurat.mouse <- scCustomize::Merge_Seurat_List(list_seurat = list("Pten" = seurat.pten, 
                                                                      "Pten_Cdcp1" = seurat.pten.cdcp1), 
                                                   add.cell.ids = c("Pten", "Pten_Cdcp1"))
    
    # create metadata for sample annotation
     seurat.mouse$sampleid <- seurat.mouse$orig.ident
     
    tail(seurat.mouse@meta.data)

  
}

```



## QC and Filter {.tabset}

### QC & filter
```{r QC_and_filter}

if(long.compute){
  
  
    # Create output dir
    output.dir <- paste0("results/QC/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    
    # set idents to sampleID
    Idents(seurat.mouse) <- seurat.mouse$sampleid
    
    # set default assay 
    DefaultAssay(seurat.mouse) <- "RNA"
    
    # Add mitochondrial and ribosomal gene percentages
    
    seurat.mouse <- Add_Mito_Ribo(object = seurat.mouse, species = "Mouse", overwrite = TRUE) 
    
    # Add cell complexity/novelty QC metric
    
   seurat.mouse <- Add_Cell_Complexity(object = seurat.mouse, overwrite = TRUE)
    
    # Setting cutoff values 
    feature.low <- 200
    feature.high <- 6000
    umi.low <- 1000
    umi.high <- 45000
    mito.high <- 10
    ribo.high <- 5
    complexity.high <- 0.8
    
    
    ########################
    # QC plots
    ########################
    if(do.plot){
      # nFeatures
      print(
        QC_Plots_Genes(seurat_object = seurat.mouse, 
                       low_cutoff = feature.low, 
                       high_cutoff = feature.high,
                       pt.size = 0.05) #how many genes per cell? See where to set the cutoffs later
      )
      
      dev.copy(pdf, paste0(output.dir, "genes.pdf"))
      dev.off()
      
      
      # UMI
      print(
        QC_Plots_UMIs(seurat_object = seurat.mouse, 
                      low_cutoff = umi.low, 
                      high_cutoff = umi.high, 
                      y_axis_log = T,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, "UMI.pdf"))
      dev.off()
      
      
      # Mito percentage
      print(
        QC_Plots_Mito(seurat_object = seurat.mouse, 
                      high_cutoff = mito.high,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, "mito.pdf"))
      dev.off()
      
      
      # Ribo percentage
      print(
        QC_Plots_Feature(seurat_object = seurat.mouse,
                         feature = "percent_ribo",
                         high_cutoff = ribo.high,
                         pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, "Ribo.pdf"))
      dev.off()
      
      
      # Complexity
      print(
        QC_Plots_Complexity(seurat_object = seurat.mouse, 
                            high_cutoff = complexity.high,
                            pt.size = 0.05)
      ) 
      
      dev.copy(pdf, paste0(output.dir, "complexity.pdf"))
      dev.off()
      
    }
    
    # Filter dataset
    seurat.mouse <- subset(seurat.mouse, 
                                  subset = nFeature_RNA > feature.low & 
                                    nFeature_RNA < feature.high & 
                                    percent_mito < mito.high & 
                                    percent_ribo > ribo.high & 
                                    log10GenesPerUMI > complexity.high)
    
    if(do.plot){
      # nFeatures filtered matrix
      print(
        QC_Plots_Genes(seurat_object = seurat.mouse, 
                       low_cutoff = feature.low, 
                       high_cutoff = feature.high,
                       pt.size = 0.05) #how many genes per cell? See where to set the cutoffs later
      )
      
      dev.copy(pdf, paste0(output.dir, "genes_filtered.pdf"))
      dev.off()
      
      
      # UMI filtered matrix
      print(
        QC_Plots_UMIs(seurat_object = seurat.mouse, 
                      low_cutoff = umi.low, 
                      high_cutoff = umi.high, 
                      y_axis_log = T,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, "UMI_filtered.pdf"))
      dev.off()
      
      # Ribo filtered percentage
      print(
        QC_Plots_Feature(seurat_object = seurat.mouse,
                         feature = "percent_ribo",
                         high_cutoff = ribo.high,
                         pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, "Ribo_filtered.pdf"))
      dev.off()
      
      
      
  
}

    
}
```


## Normalisation and integration {.tabset}

### Azimuth cell type prediction/annotation
```{r Azimuth_celltype_prediction}

do.azimuth <- FALSE
if(do.azimuth){
  
  #######################################
  # quick prediction of cell annotations
  #######################################
  
  library("Azimuth")
  DefaultAssay(seurat.mouse) <- "RNA"
  seurat.mouse <- JoinLayers(seurat.mouse)
  seurat.mouse <- RunAzimuth(seurat.mouse, reference = "pbmcref")
  
  seurat.mouse
  head(seurat.mouse@meta.data)
  
  DimPlot2(seurat.mouse, group.by = "predicted.celltype.l1", reduction = "ref.umap", label = T)
    DimPlot2(seurat.mouse, group.by = "predicted.celltype.l3", reduction = "ref.umap", label = T)

  # see available references 
  #library("SeuratData")
  #available_data <- AvailableData()
  #available_data[grep("Azimuth", available_data[, 3]), ]
  
}

```

### Normalise & integrate GEX data
```{r Normalise_and_integrate_GEX_data}

if(long.compute){
  
  # set default assay 
  DefaultAssay(seurat.mouse) <- "RNA"
  
  #####################################################
  # Normalise with scTransform and integrate data
  #####################################################
  # note error with more recent version of matrixStats resulting in "useNames = NA is defunct" error. downgrade to matrixStats version 1.1.0 as a workaround
  # https://github.com/satijalab/seurat/issues/7501#issuecomment-1854571904
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts  
  
  
  # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
  
  # Normalise and scale with SCTransform v2
  seurat.mouse <- SCTransform(seurat.mouse, 
                           method = "glmGamPoi", 
                           vars.to.regress = "percent_mito", 
                           vst.flavor = "v2",
                           verbose = TRUE) %>%
    RunPCA(reduction.name = "pca_sct") %>%
    RunUMAP(reduction = "pca_sct",
            dims = 1:30, 
            reduction.name = "umap.unintegrated.sct")
  
   DefaultAssay(seurat.mouse) <- "SCT"
  
  # integrate layers using cca integration
  seurat.mouse <- IntegrateLayers(object = seurat.mouse,
                               orig.reduction = "pca_sct",
                               new.reduction = "integrated.cca.sct",
                               method = CCAIntegration,
                               normalization.method = "SCT",
                               verbose = TRUE)
  
  # find clusters and dim reduction using integrated data
  seurat.mouse <- FindNeighbors(seurat.mouse,
                             dims = 1:30,
                             reduction = "integrated.cca.sct") %>%
    FindClusters(resolution = 1.5, 
                 cluster.name = "clusters_cca_sct") %>%
    RunUMAP(reduction = "integrated.cca.sct",
            dims = 1:30, 
            reduction.name = "umap.cca.sct")
  
  
  if(do.plot){
    
    # create output dir
    output.dir <- paste0("results/integration/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    
    unintegrated.reduction <- "umap.unintegrated.sct"
    integrated.reduction <- "umap.cca.sct"
    title.var <- "SCT"
    
    
    # evaluate integration and dim reduction
    
    vars.plot <- c("sampleid", "seurat_clusters")
    for(i in seq_along(vars.plot)){
      
      
      # remove "." from name
      name.var <- gsub("\\.", "_", vars.plot[i])
      
      before.int <- DimPlot(seurat.mouse,
                            reduction = unintegrated.reduction, 
                            label = T,
                            group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("Unintegrated ", title.var, " Normalisation"))
      
      after.int <- DimPlot(seurat.mouse,
                           reduction = integrated.reduction, 
                           label = T,
                           group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("CCA Integration using ", title.var, " Normalisation"))
      
      print(before.int | after.int)
      dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration", name.var, ".pdf"))
      dev.off()
      
      
      
    }
    
  }
  
}


```


### Overview
```{r Overview}

# create output dir
output.dir <- paste0("results/Dataset_overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

scCustomize::DimPlot_scCustom(seurat.mouse,
                              group.by = "seurat_clusters", 
                              label = T,
                              reduction = "umap.cca.sct")

dev.copy(pdf, paste0(output.dir, "UMAP_seurat_clusters.pdf"))
dev.off()



scCustomize::DimPlot_scCustom(seurat.mouse,
                              group.by = "sampleid", 
                              label = T,
                              reduction = "umap.cca.sct")

dev.copy(pdf, paste0(output.dir, "UMAP_sampleid.pdf"))
dev.off()


scCustomize::DimPlot_scCustom(seurat.mouse,
                              group.by = "predicted.celltype.l1", 
                              label = T,
                              reduction = "umap.cca.sct")

dev.copy(pdf, paste0(output.dir, "UMAP_predicted_celltype_l1.pdf"))
dev.off()



# Cluster distribution bar plot
ClusterDistrBar(seurat.mouse$sampleid, seurat.mouse$seurat_clusters)
dev.copy(pdf, paste0(output.dir, "Clust_Distribution_seurat_clusters.pdf"))
dev.off()

ClusterDistrBar(seurat.mouse$sampleid, seurat.mouse$seurat_clusters, 
                rev = T, 
                percent = T, 
                normalize = T, 
                plot = T)

dev.copy(pdf, paste0(output.dir, "Clust_Distribution_seurat_clusters_normalised.pdf"))
dev.off()

ClusterDistrBar(seurat.mouse$sampleid, seurat.mouse$seurat_clusters, 
                rev = T, 
                percent = F, 
                normalize = T, 
                plot = T)

dev.copy(pdf, paste0(output.dir, "Clust_Distribution_seurat_clusters_normalised_abs.pdf"))
dev.off()



table(seurat.mouse$sampleid, seurat.mouse$seurat_clusters)
table(seurat.mouse$sampleid)

colSums(x)



```




### Save seurat object
```{r save_seurat_object}

if(long.compute){
  
  SaveSeuratRds(seurat.mouse, file = "saves/seurat.mouse.rds")
  
  # extract and save metadata 
  write.csv(seurat.mouse@meta.data, "saves/metadata.csv")
}else{
  
  seurat.mouse <- LoadSeuratRds("savesseurat.mouse.rds")
}

```


## Annotate clusters {.tabset}

### Differential gene expression GEX vs seurat_clusters
```{r DEG_analysis_seurat_clusters}

# create output dir
output.dir <- paste0("results/DEG_analysis_seurat_clusters/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

################################################
# Caclulate GEX DEGs for across seurat_clusters 
################################################

if(long.compute){
  
  # set assay
  DefaultAssay(seurat.mouse) <- "SCT"
  seurat.mouse <- PrepSCTFindMarkers(seurat.mouse)
  
  # DEG calculation using single-cells
  
  # set assay
  DefaultAssay(seurat.mouse) <- "SCT"
  
  # format cluster annotations
  seurat.mouse@meta.data$seurat_clusters <- factor(seurat.mouse@meta.data$seurat_clusters, levels = c(0:length(unique(seurat.mouse@meta.data$seurat_clusters))))
  
  # set idents
  Idents(seurat.mouse) <- "seurat_clusters"
  
  seurat_clusters.gex.de <- FindAllMarkers(seurat.mouse,
                                           assay = "SCT",
                                           only.pos = TRUE, 
                                           verbose = TRUE)
  
  # save to file 
  write.csv(seurat_clusters.gex.de, paste0(output.dir, "DEGs_wilcox_single_cell_SCT_seurat_clusters.csv"))
  
}else{
  
  seurat_clusters.gex.de <- read.csv(paste0(output.dir, "DEGs_wilcox_single_cell_SCT_seurat_clusters.csv"))
  
}




```


### DEG heatmaps - GEX and ADT - seurat_clusters
```{r DEG_heatmaps_seurat_clusters}

# create output dir
output.dir <- paste0("results/DEG_analysis_seurat_clusters/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(do.plot){
  
  
  ############
  # GEX data
  ############
  DefaultAssay(seurat.mouse) <- "SCT"
  
  # create pseudobulk for visualisation
  pseudo.seurat <- AggregateExpression(seurat.mouse, 
                                   assays = "SCT",
                                   return.seurat = TRUE,
                                   group.by = c("seurat_clusters"))
  
  # set idents of pseudobulk object
  Idents(pseudo.seurat) <- "seurat_clusters"
  
  # extract top 10 DEGs per cluster
  goi <- seurat_clusters.gex.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::group_by(cluster) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    dplyr::pull(gene)
  
  # heatmap of top 10 degs per cluster
  DoHeatmap(pseudo.seurat, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + NoLegend()
  
  dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_seurat_clusters_pseudobulk.pdf"))
  dev.off()
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(seurat_clusters.gex.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- seurat_clusters.gex.de %>%
      dplyr::filter(p_val_adj < 0.05) %>%
      dplyr::filter(cluster == cluster.id[i]) %>%
      dplyr::top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.seurat, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE)
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk__seurat_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}

```



## not written 
### Key lineage markers
```{r key_lineage_markers}

# create output dir
output.dir <- paste0("results/key_lineage_markers/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



goi <- c("Cd3e", "Cd20", "Ncam1", "Foxp3", "Cd8a", "Cd8b", "Cd4", "Nkg7")

DoHeatmap(pseudo.seurat, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + NoLegend()
  
#dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_seurat_clusters_pseudobulk.pdf"))
 # dev.off()






ggplot2_switch_version(desired_version = "3.4.4")


# set up object
DefaultAssay(HL.seurat) <- "ADT"

goi <- rownames(HL.seurat@assays$ADT)


# Plot Feature and Density plots iteratively over goi 
for(i in seq_along(goi)){
  
  # feature plots
  print(
    scCustomize::FeaturePlot_scCustom(HL.seurat, 
                                      reduction = "wnn.umap", 
                                      features = goi[i])
    
  )
  
  dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_", goi[i], ".pdf"))
  dev.off()
  
  
  # density plots
  
  print(
    scCustomize::Plot_Density_Custom(HL.seurat, 
                                     reduction = "wnn.umap",
                                     features = goi[i], 
                                     custom_palette = batlow.pal)
  )
  dev.copy(pdf, paste0(output.dir, "UMAP_DensityPlot_", goi[i], ".pdf"))
  dev.off()
  
  
  
}


ggplot2_switch_version(desired_version = "3.5.0")


```





## Below code kept for reference purposes




## Doublet detection {.tabset}

### Preprocess seurat object for doublet detection
```{r preprocess_for_doublet_detection}

if(long.compute){
  #################################################################
  # For doublet detection normalise, and perform dim reduction
  #################################################################
  
  # set default assay 
  DefaultAssay(HL.seurat) <- "RNA"
  
  # split object up based on sequencing run / in seurat V5 these are splitting the layers
  HL.seurat[["RNA"]] <- split(HL.seurat[["RNA"]],  f = HL.seurat$RunID)
  
  # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
  
  # Normalise and scale with SCTransform v2
  HL.seurat <- SCTransform(HL.seurat, 
                           method = "glmGamPoi", 
                           vars.to.regress = "percent_mito", 
                           vst.flavor = "v2",
                           verbose = TRUE) %>%
    RunPCA(reduction.name = "pca_sct") %>%
    RunUMAP(reduction = "pca_sct",
            dims = 1:30, 
            reduction.name = "umap.unintegrated.sct")
  
  # quick vis of unintegrated data 
  DimPlot_scCustom(HL.seurat, 
                   reduction = "umap.unintegrated.sct", 
                   pt.size = 1, 
                   group.by = "RunID")
  
  
  DefaultAssay(HL.seurat) <- "SCT"
}

```

### Use DoubletFinder to identify and then remove doublets
```{r DoubletFinder}

if(long.compute){
  
  #remotes::install_github('chris-mcginnis-ucsf/DoubletFinder')
  library("DoubletFinder")
  
  HL.seurat <- SplitObject(HL.seurat, split.by = "RunID")
  
  # unfortunately keep hitting errors with parallelized code execution so convert to sequential mode 
  plan("sequential")
  
  
  
  ## pK Identification (no ground-truth) 
  #sweep.res.list <- lapply(HL.seurat, 
  #                         FUN = paramSweep, 
  #                         PCs = 1:30, 
  #                         sct = TRUE, 
  #                         num.cores = ncores)
  
  #sweep.stats <- lapply(sweep.res.list, 
  #                     FUN = summarizeSweep,
  #                    GT = FALSE)
  
  #bcmvn <- lapply(sweep.stats, FUN = find.pK)
  
  #bcmvn$Run_1 %>%
  #  dplyr::top_n(5, MeanBC)
  
  ## Homotypic Doublet Proportion Estimate # use run3 which has the most cells
  nExp_poi <- round(0.075*nrow(HL.seurat$Run_3@meta.data))  ## Assuming 7.5% doublet formation rate - tailor for your dataset
  
  ## Run DoubletFinder with varying classification stringencies ----------------------------------------------------------------
  #HL.seurat <- lapply(HL.seurat, 
  #                    FUN = doubletFinder,
  #                    PCs = 1:30, 
  #                    pN = 0.25, 
  #                    pK = 0.0005, 
  #                    nExp = nExp_poi,
  #                    reuse.pANN = FALSE, 
  #                    sct = TRUE)
  
  #HL.seurat <- lapply(HL.seurat, 
  #                    FUN = doubletFinder,
  #                    PCs = 1:30, 
  #                    pN = 0.25, 
  #                    pK = 0.09, 
  #                    nExp = nExp_poi,
  #                    reuse.pANN = FALSE, 
  #                    sct = TRUE)
  
  HL.seurat <- lapply(HL.seurat, 
                      FUN = doubletFinder,
                      PCs = 1:30, 
                      pN = 0.25, 
                      pK = 0.3, 
                      nExp = nExp_poi,
                      reuse.pANN = FALSE, 
                      sct = TRUE)
  
  # move back to a parallized mode
  plan("multisession", workers = ncores) #multisession
  
  # quick look at Run3 as this has the most cells
  #table(HL.seurat$Run_3@meta.data$`DF.classifications_0.25_5e-04_1440`)
  #table(HL.seurat$Run_3@meta.data$DF.classifications_0.25_0.09_1440)
  #table(HL.seurat$Run_3@meta.data$DF.classifications_0.25_0.3_1440)
  
  # manually looked through all runs and no difference between params used
  # as in exact same number of singlets/doublets regardless of params, true for all runs 
  
  
  ################################################
  # Merge sequencing Runs into single object
  ################################################
  
  # some error with the ADT assay preventing merging, renaming the assay seems to fix this problem
  # Rename the 'ADT' assay in one of the Seurat objects
  HL.seurat <- lapply(HL.seurat, 
                      FUN = RenameAssays, 
                      assay.name = "ADT",
                      new.assay.name = "ADT_2")
  
  
  HL.seurat <- scCustomize::Merge_Seurat_List(HL.seurat, 
                                              merge.data = TRUE)
  
  # ensure layers are all joined
  HL.seurat <- JoinLayers(HL.seurat,  assay = "RNA")
  HL.seurat <- JoinLayers(HL.seurat,  assay = "ADT_2")
  HL.seurat <- JoinLayers(HL.seurat,  assay = "Hashtag")
  
  
  # rename the ADT assay back to default 
  HL.seurat <- RenameAssays(HL.seurat,
                            assay.name = "ADT_2", 
                            new.assay.name = "ADT")
  
  # cell names got a leading "_" appended to them for example "_Run_1_ACGCCAGTCTCGCATC-1" 
  # remove the leading "_" 
  cell.names <- Cells(HL.seurat)
  new.cell.names <- gsub("^_", "", cell.names)
  HL.seurat <- RenameCells(HL.seurat, new.names = new.cell.names)
  
  
  ################################################
  # Remove doublets 
  ################################################
  logic.vec <- grepl("DF.classifications_0.25_0.3_.*", colnames(HL.seurat@meta.data))
  
  
  table(HL.seurat@meta.data[,logic.vec]) # 48,790 singlets and 5,760 doublets 
  
  
  
  # distribution of doublets across hash status
  table(HL.seurat@meta.data$HashStatus, HL.seurat@meta.data[,logic.vec])
  
  # cols = doubletfinder
  # rows = hashtag results
  #           Doublet Singlet
  #  Negative    1003    8203
  #  Singlet     4757   40587
  
  # look at doublets across donor, look to be relatively evenly distributed
  x <- table(HL.seurat@meta.data$DonorID, HL.seurat@meta.data[,logic.vec])
  clust.val <- rowSums(x) 
  freq.vals <- sweep(x, MARGIN = 1, clust.val, FUN = "/")
  freq.vals <- freq.vals*100
  freq.vals
  
  # Patient 8 has a high 36% doublet rate
  
  # remove doublets 
  Idents(HL.seurat) <- HL.seurat@meta.data[,logic.vec]
  
  # initial values 
  # 60133 features across 54550 cells
  HL.seurat
  HL.seurat <- subset(HL.seurat, idents = "Singlet")
  
  HL.seurat # 48790 cells across 60,133 genes
  # thefore lost 5,760 cells
  
  table(HL.seurat@meta.data$HashStatus) # 8,203 cells are negative for having a Hashtag
  
  
}


```


### Save seurat object
```{r save_seurat_object}

if(long.compute){
  
  SaveSeuratRds(HL.seurat, file = "saves/HL_seurat_Doubletfilt.rds")
  
  # extract and save metadata 
  write.csv(HL.seurat@meta.data, "saves/metadata.csv")
}else{
  
  HL.seurat <- LoadSeuratRds("saves/HL_seurat_Doubletfilt.rds")
}

```



### Save seurat object
```{r save_seurat_object}

if(long.compute){
  
  SaveSeuratRds(HL.seurat, file = "saves/HL_seurat_Azimuth.rds")
  
  # extract and save metadata 
  write.csv(HL.seurat@meta.data, "saves/metadata.csv")
}else{
  
  HL.seurat <- LoadSeuratRds("saves/HL_seurat_Azimuth.rds")
}

```



## Cluster identification and UMAP optimisation {.tabset}


### Cluster optimisation
```{r Cluster_optimisation}


if(long.compute){
  
  
  # create output dir
  output.dir <- paste0("results/Cluster_and_UMAP_optimisation/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  Graphs(HL.seurat, slot = NULL)
  Neighbors(HL.seurat, slot = NULL)
  Reductions(HL.seurat, slot = NULL)
  
  # Set seed for reproducibility
  set.seed(42)
  
  
  
  
  DefaultAssay(HL.seurat) <- "SCT"
  
  #neihbor.name <- "wknn"
  graph.name <- "wsnn"
  umap.reduction <- "wnn.umap"
  prefix.var <- "wsnn_res."
  title.var <- "wnn"
  
  
  
  
  # Find clusters over an iteration of resolutions
  HL.seurat <- Seurat::FindClusters(HL.seurat, 
                                    resolution = seq(0, 2, by = 0.1), 
                                    graph.name = graph.name)  
  
  
  head(HL.seurat@meta.data)
  
  # Use clustree package to visualise impact of cluster resolution on clusters determined
  
  #install.packages("clustree")
  
  ggplot2_switch_version(desired_version = "3.5.0")
  
  library(clustree) # clusttree needs to be loaded, error occurs if just call function from namespace
  
  clustree::clustree(HL.seurat, 
                     prefix = prefix.var,
                     highlight_core = TRUE,
                     layout = "tree") #tree sugiyama
  
  dev.copy(pdf, paste0(output.dir, "clusttree_wnn_tree.pdf"))
  dev.off()
  
  
  # quick iterate over diff clust resolutions
  res.vals <- grep("wsnn_res.*", colnames(HL.seurat@meta.data), value = T)
  
  for(i in seq_along(res.vals)){
  print(DimPlot(HL.seurat,
                reduction = "wnn.umap", 
                label = T,
                group.by = res.vals[i]) + 
          NoLegend() + 
          ggtitle(paste0(res.vals[i]))
  )
  
  name.var <- gsub("\\.", "_", res.vals[i])
  
  dev.copy(pdf, paste0(output.dir, "wnnUMAP_", name.var, ".pdf"))
  dev.off()
  }
  
  # based on above plots, 0.9 looks best where stability is reached but not over clustering

  # set seurat clusters to this level 
  HL.seurat@meta.data$seurat_clusters <- HL.seurat@meta.data$wsnn_res.0.9

  length(unique(HL.seurat@meta.data$seurat_clusters)) # 25 clusters
  
  
  
  
  # Visualise clusters on UMAP projection 
  DimPlot(HL.seurat,
          reduction = "wnn.umap", 
          label = T,
          group.by = "seurat_clusters") + 
    NoLegend() + 
    ggtitle("wsnn_res.0.9")
  
  dev.copy(pdf, paste0(output.dir, "wnnUMAP_seurat_clusters.pdf"))
  dev.off()
  
  # some clusters will need to be aggregated
  
  
  # Optimise UMAP projection 
  
  # set init conditions and step sizes
  init.min.dist = 0.001
  init.neigh = 5
  min.dist.step = 2 
  neigh.step = 20
  
  min.dist.val <- init.min.dist
  
  # start loop 
  while (min.dist.val < 1) {
    
    n.neigh.val <- init.neigh
    
    while (n.neigh.val <= 50) {
      
      print(paste0("Calculating UMAP for min.dist = ", 
                   min.dist.val, " & n.neighbor = ", n.neigh.val))
      
      
      
      # caculate UMAP
      temp <- Seurat::RunUMAP(object = HL.seurat,
                              nn.name = "weighted.nn", 
                              n.neighbors = n.neigh.val, 
                              min.dist = min.dist.val,
                              reduction.name = "umap.temp",
                              seed.use = 42)
      
      # print UMAP
      p1 <- Seurat::DimPlot(object = temp, 
                            reduction = "umap.temp", 
                            label = TRUE, 
                            label.size = 4, 
                            group.by = "seurat_clusters") + 
        ggtitle(paste0("Seurat_clusters - UMAP Min dist = ", 
                       min.dist.val, " n.val = ", n.neigh.val)) +
        NoLegend()
      
      p2 <- Seurat::DimPlot(object = temp, 
                            reduction = "umap.temp", 
                            label = TRUE, 
                            label.size = 4, 
                            group.by = "predicted.celltype.l2") + 
        ggtitle("predicted.celltype.l2") +
        NoLegend()
      
      print(p1|p2)
      
      dev.copy(pdf, 
               paste0(output.dir, 
                      "UMAP_Min_dist_", min.dist.val, "_neighval_", n.neigh.val, ".pdf"))
      dev.off()
      
      n.neigh.val <- n.neigh.val + neigh.step
    }
    min.dist.val <- min.dist.val * min.dist.step
  }
  
  # structure is rather stable with very little difference, will use default params
  
  
  # calculate UMAP using default params 
  
  HL.seurat <- RunUMAP(HL.seurat, 
                       nn.name = "weighted.nn", 
                       reduction.name = "wnn.umap", 
                       reduction.key = "wnnUMAP_")
  
  
}


```


### Save seurat object
```{r save_seurat_object}

if(long.compute){
  
  SaveSeuratRds(HL.seurat, file = "saves/HL_seurat_integrated.rds")
  
  # extract and save metadata 
  write.csv(HL.seurat@meta.data, "saves/metadata.csv")
}else{
  
  HL.seurat <- LoadSeuratRds("saves/HL_seurat_integrated.rds")
}

```




